Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-05-01 manuscript/manuscript.tex
discuss centrality + address last of fabiennes comments

\documentclass[twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering
\usepackage[shortcuts]{extdash} 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, \\
GENESIS Study collaborators, Dominique Stoppa-Lyonnet$^{4,5,6}$, \\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women with a family history. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, a ubiquitin ligase related that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that most of the genome contributes to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Additionally, as genes that contribute to a disease tend to participate in similar biological functions, studying the neighborhood of known disease genes is effective at identifying new ones \cite{huang_systematic_2018}.

@@

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for \emph{BRCA1} and \emph{BRCA2} pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of their corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as functional candidate variants in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P\=/value \textless~0.001), and those missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After controlling for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, 2 cases being related to an index case, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1/2} mutation detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could produce spurious associations. A PCA revealed no visual differential population structure between cases and controls (Supplementary Figure~\ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\)~=~1.05) either, further confirming the absence of confounding population structure.

@@

\paragraph{Gene network}
The statistical frameworks of the different network methods are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.). Yet, we used protein-protein interaction networks (PPIN) for all of them except SConES, as they are interpretable, well characterized, and they were designed to scale appropriately to them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols, involving 13\,619 genes. 12\,880 of those mapped to a genotyped SNP after quality control, involving 127\,604 interactions. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only network method designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}). The properties of these three subnetworks are available in Supplementary Table \ref{tab:snp_solutions}.

@@

\begin{table}[htbp]
  \caption{Summary of the differences between the network methods.}
  \label{tab:method_comparison}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & \(\ge\) 1 & Summary & $\Phi$\textsuperscript{-1}(1 - P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of nodes in the network, either genes (PPIN) or SNPs. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than containing high scoring, connected nodes. \textbf{\emph{Comp.}}: number of connected components in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P\=/values are transformed into node scores. In the case of SigMod, $\Phi$\textsuperscript{-1} represents the inverse of the
        cumulative distribution function of the standard Normal distribution. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subgraph on the network \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P\=/value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P\=/values by a beta-uniform model (BUM) parameterized by the desired false discovery rate (FDR). Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitively long computational time. To solve it efficiently it is re-cast as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P\=/value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterwards.

@@

We evaluated multiple properties of the different methods (described below) through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures.

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 13\,250\,642 SNPs (after imputation) genotyped on 228\,951 women of European ancestry from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic, while GENESIS is a homogeneous dataset focusing on the French high-risk population attending the family cancer clinics. Despite these differences, we expect some degree of shared genetic architecture, especially at the gene level. For that purpose, we searched associated genes as in Section~\ref{methods:node_score}. We used all the SNPs to measure gene association, but only those available after quality control on GENESIS for the SNP-level comparisons.

@@

\subsection{A conventional GWAS confirms that FGFR2 is associated with familial breast cancer}
\label{results:conventional}

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P\=/value lower than the Bonferroni threshold in chromosomes 10 and 16 (Supplementary Figure~\ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} were related to breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1/2} carriers studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for \emph{BRCA1/2} \cite{rinella_genetic_2013}. Only \emph{FGFR2} was significant at the gene-level (Supplementary Figure~\ref{sfig:snp_gene_manhattan}B). These results show an overlap in the genetic architecture of the disease between the studied French population sample and other populations.

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and from other regions (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Figure~\ref{fig:benchmark}B), while providing more interpretable solutions.

@@

\subsection{A case study: the consensus network}
\label{results:consensus}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{./figures/figure_3.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significant in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Supplementary Figure \ref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

@@

\subsection{The consensus overcomes the limitations of the individual solutions}
\label{results:drawbacks}

>>>>

\documentclass[draft,twocolumn, 11pt]{article}
\usepackage{amsmath,amssymb}
\usepackage[format=hang,labelfont=bf,textfont=small,singlelinecheck=false,justification=justified,margin={20pt,20pt},figurename=Fig.]{caption}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{amssymb}
\usepackage{capt-of}
\usepackage{hyperref}
\usepackage{threeparttable}
\usepackage[numbers]{natbib}
\usepackage{lmodern} % Chloe: better font rendering
\usepackage[shortcuts]{extdash} 
\newcommand{\mean}[1]{$\overline{\mbox{#1}}$}
\newcommand{\median}[1]{$\hat{\mbox{#1}}$}

@@

\title{Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer}
\author{Héctor Climente-González$^{1,2,3}$, Christine Lonjou$^{1,2,3}$, Fabienne Lesueur$^{1,2,3}$, \\
GENESIS Study collaborators$^{1,2,3}$, Dominique Stoppa-Lyonnet$^{4,5,6}$, \\
Nadine Andrieu$^{1,2,3}$, Chloé-Agathe Azencott$^{3,1,2}$}
\date{$^{1}$Institut Curie, PSL Research University, F-75005 Paris, France;\\
  $^{2}$INSERM, U900, F-75005 Paris, France;\\
  $^{3}$MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
  $^{4}$Service de Génétique, Institut Curie, F-75005 Paris, France;\\
  $^{5}$INSERM, U830, F-75005 Paris, France;\\
  $^{6}$Université Paris Descartes.\\
}
\hypersetup{
  pdfauthor={Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, GENESIS Study collaborators, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott},
  pdftitle={Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer},
  pdfkeywords={},
  pdfsubject={},
  pdflang={English}}
\begin{document}

@@

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study six network methods which identify subnetworks with high association scores to a phenotype. Specifically, we examine their utility to discover new biomarkers for breast cancer susceptibility by interrogating a genome-wide association study (GWAS) focused on French women with a family history of breast cancer and tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. For instance, we show a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Importantly, we find a significantly large overlap between the genes in the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Yet, network methods are notably unstable, producing different results when the input data changes slightly. To account for that, we produce a stable consensus subnetwork, formed by the most consistently selected genes. The stable consensus is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than average. The network seems organized around \emph{CUL3}, encoding an ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are not well understood. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are very functionally inter-related and influence each other's behavior, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might address the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}. This is particularly true for cancer driver genes, which tend to be key players in connecting different, densely-connected communities of genes. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

@@

The GENE Sisters (GENESIS) study was designed to investigate risk factors for familial breast cancer in the French population \cite{sinilnikova_genesis:_2016}. Index cases are patients with infiltrating mammary or ductal adenocarcinoma, who had a sister with breast cancer, and who have been tested negative for \emph{BRCA1} and \emph{BRCA2} pathogenic variants. Controls are unaffected colleagues and/or friends of the cases, born around the year of birth of their corresponding case (\textpm{} 3 years). We focused on the 2\,577 samples of European ancestry, of which 1\,279 are controls and 1\,298 are cases. The genotyping was performed using the iCOGS array, a custom Illumina array designed to study genetic susceptibility of hormone-related cancers \cite{sakoda_turning_2013}. It contains 211\,155 SNPs, including SNPs putatively associated with breast, ovarian, and prostate cancers, SNPs associated with survival after diagnosis, and SNPs associated to other cancer-related traits, as well as candidate functional variants in selected genes and pathways.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P\=/value \textless~0.001), and those with genotyping data missing on more than 10\% of the samples. A subset of 20 duplicated SNPs in \emph{FGFR2} were also removed. In addition, we removed the samples with more than 10\% missing genotypes. After controlling for relatedness, 17 additional samples were removed (6 for sample identity error, 6 controls related to other samples, 2 cases being related to an index case, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1} or \emph{BRCA2} pathogenic variant detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

We looked for population structure that could produce spurious associations. A principal component analysis revealed no visual differential population structure between cases and controls (Supplementary Figure~\ref{sfig:pcs}). Independently, we did not find evidence of genomic inflation (\(\lambda\)~=~1.05) either, further confirming the absence of confounding population structure.

@@

\paragraph{Gene network}
The statistical frameworks of the different network methods are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.). Yet, we used protein-protein interaction networks (PPIN) for all of them except SConES, as they are interpretable, well characterized, and they were designed to run efficiently on networks of their size. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols, involving 13\,619 genes. 12\,880 of those mapped to a genotyped SNP after quality control, involving 127\,604 interactions. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only network method designed to handle SNP networks. As in gene networks, two SNPs are connected in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}). The properties of these three subnetworks are available in Supplementary Table \ref{tab:snp_solutions}.

@@

\begin{table}[htbp]
  \caption{Summary of the differences between the network methods.}
  \label{tab:method_comparison}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & \(\ge\) 1 & Summary & $\Phi$\textsuperscript{-1}(1 - P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of nodes in the network, either genes (PPIN) or SNPs. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than containing high scoring, connected nodes. \textbf{\emph{Comp.}}: number of connected components in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P\=/values are transformed into node scores. In the case of heinz, BUM stands for beta-uniform model, used to transform the P-values; for SigMod, $\Phi$\textsuperscript{-1} represents the inverse of the
        cumulative distribution function of the standard Normal distribution. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\begin{description}
\item[{heinz}] The goal of heinz is to identify the highest-scored connected subnetwork \cite{dittrich_identifying_2008}. The authors propose a transformation of the genes' P\=/value into a score that is negative under no association with the phenotype, and positive when there is. This transformation is achieved by modelling the distribution of P\=/values by a beta-uniform model (BUM) parameterized by the desired false discovery rate (FDR). Thus formulated, the problem is NP-complete, and hence solving it would require a prohibitively long computational time. To solve it efficiently it is re-cast as the Prize-Collecting Steiner Tree Problem (PCST), which seeks to select the connected subnetwork S that maximizes the \emph{profit} p(S), defined as:

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). We adjusted the P-values for multiple testing as in \citet{Benjamini1995} (BH).Pathways with an BH adjusted P\=/value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterwards.

@@

We evaluated multiple properties (described below) of the different methods through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures.

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries which included 13\,250\,642 SNPs genotyped or imputed on 228\,951 women of European ancestry mostly from the general population \cite{Michailidou2017}. Hence, a high proportion of breast cancer cases investigated in BCAC are sporadic (not selected according to family history), while GENESIS is a homogeneous dataset not included in BCAC and which focus on the French high-risk population attending the family cancer clinics. Despite these differences, we expect some degree of shared genetic architecture, especially at the gene level. For that purpose, we searched associated genes as in Section~\ref{methods:node_score}. We provided VEGAS2 with both the summary statistics of all available SNPs, and the genotypes from European samples from the 1000 Genomes Project \cite{the_1000_genomes_project_consortium_global_2015} to compute the LD patterns.

@@

\subsection{Conventional SNP- and gene-based analyses confirm that \emph{FGFR2} locus is associated with familial breast cancer}
\label{results:conventional}

We conducted association analyses in the GENESIS dataset at both SNP and gene levels (Section~\ref{methods:node_score}). Two genomic regions have a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (Supplementary Figure~\ref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16}, and it is located near the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In our studied population, only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (Supplementary Figure~\ref{sfig:snp_gene_manhattan}B).

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and new ones (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Figure~\ref{fig:benchmark}B), while providing more interpretable solutions.

@@

\subsection{A case study: the consensus network}
\label{results:consensus}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=.5\linewidth]{./figures/figure_3.pdf}
  \caption{Consensus subnetwork on GENESIS (Section~\ref{methods:methods}). Each node is represented by a pie chart, which shows the methods that selected it. We labeled (and enlarged) the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset. The latter ones are also colored in pink. All gene names are indicated in Supplementary Figure \ref{sfig:consensus-names}.}
  \label{fig:consensus}
\end{figure}

@@

\subsection{Network topology matters, and might lead to ambiguous results}
\label{results:drawbacks}
